### 2020-03-26 thesis/content/chapter-graph-lta.tex
reorganized thesis: wl2gnn gets its own chapter
includes the intro for the wl2gcnn chapter
+ various improvements

In the previous chapter an introduction to two separate fields of research was given:
\begin{enumerate*}
	\item \Acf{lta},
	\item \Acf{gcr}
\end{enumerate*}.
In this chapter we will combine them and define an extension of \ac{lta} to the \ac{gcr} problem.
This will be done in three steps:
\begin{enumerate}
	\item We begin with a formal definition of what actually constitutes an \ac{lta} method as opposed to non-\acs{lta} methods.
	\item Using this definition, we will see that some of the previously described \ac{gcr} methods can be interpreted as \ac{lta} variants under certain conditions.
	\item Finally a novel \acs{lta}-inspired \ac{gcnn} architecture will be described.
\end{enumerate}

@@

\section{\acs*{lta} Formulations of Existing \acs*{gcr} Methods}%
\label{sec:ltag:formulation}

Based on the general definition of \ac{lta} from the last section, we will now see to what extent the \ac{gcr} methods described in \cref{sec:related:gcr} can be interpreted as \ac{lta} instances.
First the relation between \acp{svm} on graph embeddings and \ac{lta} will be explored.
Then we will provide an \ac{lta} perspective on \acp{gcnn}.

\subsection{\acsp*{svm} using Graph Embeddings as \acs*{lta} Models}%
\label{sec:ltag:formulation:svm}

In \cref{sec:related:gcr:embed,sec:related:gcr:kernel} three different ways to map a given graph $G$ to a vector $\varphi(G) \in \mathbb{R}^d$ were described:
\begin{enumerate*}
	\item Fingerprint embeddings,
	\item skip-gram inspired embeddings and
	\item kernel embeddings
\end{enumerate*}.
One common way to solve the \ac{gcr} problem via those embedding vectors is to train an \ac{svm} on them.
We will now see that \acp{svm} can be interpreted as \ac{lta} models if they are trained on so-called \acp*{sce}.
\begin{defn}
	Given a a multiset $A = \ldblbrace \underbrace{a_1, \dots, a_1}_{\gamma_A(a_1)\text{ times}},\allowbreak \dots, \underbrace{a_n, \dots, a_n}_{\gamma_A(a_n)\text{ times}} \rdblbrace \subseteq D$, the so-called \textit{multiplicity function} $\gamma_A: D \to \mathbb{N}_0$ maps each element of the domain $D$ to its multiplicity in $A$ with $\gamma_A(x) = 0 \Leftrightarrow x \notin A$.
\end{defn}
\begin{defn}\label{defn:ltag:substruct-embedding}
	A graph embedding $\varphi: \mathcal{G} \to \mathbb{N}_0^{d}$ is called a \ac{sce} iff.\ %
	there are decomposition functions $\psi_{\varphi, i}: \mathcal{G} \to \mathcal{P}(\mathcal{G})$ and multiplicity functions $\gamma_{\varphi, i}: \mathcal{G} \to \mathbb{N}_0$ for all embedding components $i \in [d]$ s.t.\ %
	$\forall G \in \mathcal{G}, i \in [d]: \varphi(G)[i] = \sum_{c \in \psi_{\varphi, i}(G)} \gamma_{\varphi, i}(c)$ where $\gamma_{\varphi, i}$ is the multiplicity function of a multiset of the constituents $\psi_{\varphi, i}(G)$.
	We call $\psi_{\varphi, i}$ an \textit{underlying decomposition} of the $i$-th component of the embedding $\varphi$.
	Similarly the joint decomposition $\psi_{\varphi}(G) \coloneqq \bigcup_{i=1}^d \psi_{\varphi,i}(G)$ is an \textit{underlying decomposition} of $\varphi$.
\end{defn}
Intuitively \cref{defn:ltag:substruct-embedding} states that the value of each \ac{sce} component must be derived from the number of constituents produced by some decomposition function where each constituent might be counted multiple times.
Based on this requirement we now proof the main theorem which shows the relation between \acp{svm} and \ac{lta}.

@@

\subsection{\acsp*{gcnn} as \acs*{lta} Models}%
\label{sec:ltag:formulation:gcnn}

@@

Looking back, an overview of graph convolution and graph pooling layers was given in \cref{sec:related:gcr:nn}.
There are many ways to combine those layers in a concrete \ac{gnn} but we focus on \textit{global pooling \ac{gcnn} architectures} that use the following three steps:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Convolution:}
		First a stack of $T$ convolution layers is applied to the input feature matrix $X \in \mathbb{R}^{n \times d^{(0)}}$ where each row is a feature vector $x_i \in \mathbb{R}^{d^{(0)}}$ for the vertex $v_i \in \mathcal{V}_G$ or the vertex $k$-multiset $s_i \subseteq \mathcal{V}_G$ in case of $k$-\acsp{gnn}.
		The convolution layers produce a convolved feature matrix $Z \in \mathbb{R}^{n \times d^{(T)}}$ where each row is a convolved feature vector $z_i \in \mathbb{R}^{d^{(\text{pool})}}$.
	\item \textbf{Pooling:}
		Then the convolved feature matrix $Z$ is reduced to a single graph feature vector $z_G \in \mathbb{R}^{d^{(\text{pool})}}$ via a pooling layer.
	\item \textbf{\ac{mlp}:}
		Lastly the final output $y_G \in \mathcal{Y}$ is computed by applying a standard \ac{mlp} to the graph feature vector $z_G$.
\end{enumerate}
\[
	\text{\acs*{gcnn}}(G) \coloneqq \mathit{\acs*{mlp}}(\mathit{Pool}(\mathit{Conv}(G)))
\]
\Cref{fig:ltag:gcnn-structure} illustrates this three step architecture for vertex neighborhood convolutions like those used by \acp{gcn}~\cite{Kipf2017} or \acp{gin}~\cite{Xu2018}.
Intuitively the convolution layers implicitly define an \ac{lta} decomposition function $\psi$ and a local evaluation function $f$ while the pooling layer roughly corresponds to an \ac{lta} aggregation function $\mathcal{A}$.
This intuition is now formalized.
\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{gfx/graph-lta/gcnn-structure.pdf}
	\caption[Computational steps of a \ac{gcnn} that uses a single vertex neighborhood convolution layer.]{
		Computational steps of a \ac{gcnn} that uses a single vertex neighborhood convolution layer.
		The colored \ac{bfs} trees next to each convolved feature vector $z_i$ show the vertices $v_j$ whose input feature vectors $x_j$ were used to compute $z_i$.
	}\label{fig:ltag:gcnn-structure}
\end{figure}

@@

Based on the decomposition function $\psi: \mathcal{G} \to \mathcal{P}(\mathcal{G})$ and the multi-score evaluation function $f^{*}: \mathcal{G} \to {(\mathcal{Y} \times \mathbb{R}_{\geq 0})}^{*}$ which are implicitly defined by the stack of convolution layers $\mathit{Conv}$ of a given \ac{gcnn}, we now complete our \ac{gcnn} \ac{lta} formulation by defining an evaluation function $f: \mathcal{G} \to \mathcal{Y} \times \mathbb{R}_{\geq 0}$ as well as an aggregation function $\mathcal{A}: {(\mathcal{Y} \times \mathbb{R}_{\geq 0})}^{*} \to \mathcal{Y}$ based on the pooling layer $\mathit{Pool}: \mathbb{R}^{n \times d^{(T)}} \to \mathbb{R}^{d^{(\text{pool})}}$.

@@

	A pooling layer $\mathit{Pool}: \mathbb{R}^{n \times d^{(T)}} \to \mathbb{R}^{d^{(\text{pool})}}$ is called an \textit{associative weighted aggregation layer} iff.

@@

		\item there is a homomorphism $(\tau_{1}: \mathbb{R}^{d^{(T)}} \to (\mathcal{Y} \times \mathbb{R}_{\geq 0}), \tau_{2}: \mathbb{R}^{d^{(\text{pool})}} \to \mathcal{Y})$ from the associative algebra of $\mathit{Pool}$ to that of $\mathcal{A}$

@@

	A \ac{gcnn} of the form $\mathit{Pool}(\mathit{Conv}(G))$ has a nontrivial \ac{lta} formulation if $\mathit{Conv}$ is a stack of neighborhood convolution layers and $\mathit{Pool}$ is an associative weighted aggregation layer.

@@

		The \ac{lta} formulation in this proof assumes that the \ac{gcnn}'s output is translated into the target score space via $\tau_{2}: \mathbb{R}^{d} \to \mathcal{Y}$.
		All this means is that an \acs{lta}-like \ac{gcnn} must produce outputs $y \in \mathcal{Y}$;
		for example if a \ac{gcnn} returns one-hot vector encodings of discrete classes those must be decoded.

@@

The reason for this is that the composition $\mathit{\acs*{mlp}} \circ \mathit{Pool}$ cannot be guaranteed to be homomorphic to a weighted aggregation function $\mathcal{A}$ due to the universal approximation theorem.

To conclude our analysis of the relation between \ac{lta} and \acp{gcnn}, we check which of the graph pooling layers described in \cref{sec:related:gcr:nn} satisfy \cref{defn:ltag:assoc-aggregation-pooling} are therefore allowed in \acs{lta}-like \acp{gcnn} by \cref{thm:ltag:gcnn-ltag-formulation}.
For simplicity we assume that the target score domain is $\mathcal{Y} = [0, 1]$.
\begin{enumerate}[label={\textbf{\arabic*.}}]
	\item \textbf{Unweighted mean, minimum and maximum pooling:}
		If the last convolution layer has a single output dimension and an activation function with the signature $\sigma: \mathbb{R} \to [0,1]$ (e.g.\ the logistic function), the simple unweighted pooling layers $\mean$, $\min$ and $\max$ trivially satisfy \cref{defn:ltag:assoc-aggregation-pooling} due to their associativity.
	\item \textbf{SortPooling \& \ac{sagpool}:}
		Apart from such unweighted pooling layers we also looked at two state-of-the-art graph pooling layers:
		\begin{enumerate*}[label={\arabic*.}]
			\item SortPooling~\cite{Zhang2018} works by aggregating the top-$p$ convolved vectors $z_i \in \mathbb{R}^{d^{(T)}}$ based on a lexicographic ordering,
			\item \ac{sagpool}~\cite{Lee2019} also aggregates the top-$p$ vectors $z_i$ but determines their order based on attention weights which are encoded as one component of each $z_i$
		\end{enumerate*}.
		Both of those pooling approaches are not compatible with the \ac{lta} formulation from \cref{thm:ltag:gcnn-ltag-formulation} because top-$p$ aggregation is generally not associative.
		This is shown by the fact that there is no $\rho: \mathbb{R}^{* \times d^{(T)}} \to \mathbb{R}^{ d^{(T)}}$ s.t.\ for all vectors $z_1 < z_2$ and and aggregation functions $\mathcal{A}$ it holds that%
		\begin{alignat*}{2}
			&& \mathcal{A}(\mathit{top}_{2}(\ldblbrace z_1, z_2, z_2 \rdblbrace)) &= \mathcal{A}(\mathit{top}_{2}(\ldblbrace z_1, \rho(\ldblbrace z_2, z_2 \rdblbrace) \rdblbrace)) \\
			\Leftrightarrow&& \mathcal{A}(\ldblbrace z_2, z_2 \rdblbrace) &= \mathcal{A}(\ldblbrace z_1, \rho(\ldblbrace z_2, z_2 \rdblbrace) \rdblbrace)\quad\lightning
		\end{alignat*}
		From the \ac{lta} perspective the non-associativity of SortPooling and \ac{sagpool} implies that, if they were used in a \ac{gcnn}, it could happen that some subtree evaluations $(y_i, w_i) \in f^{*}(c)$ of a single constituent $c$ are considered by the aggregation function $\mathcal{A}$ while some are ignored depending on the subtree evaluations of other constituents, i.e.\ the constituent evaluation $f(c)$ could then depend on other constituents.
		This is forbidden in our definition of \ac{lta}.
\end{enumerate}

Since neither SortPooling nor \ac{sagpool} are compatible with \ac{lta}, we propose an associative weighted aggregation layer based on \ac{sagpool} which uses a softmax-based attention filter instead of a top-$p$ filter.
\begin{defn}
	We define \ac{sampool} as
	\[\mathit{SAM}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) \coloneqq\frac{1}{\sum_{i=1}^n e^{w_i}} \sum_{i=1}^n e^{w_i} y_i \text{.}\]
\end{defn}
If the stack of convolution layers of a \ac{gcnn} produces vectors $z_i = (y_i, w_i) \in [0, 1] \times \mathbb{R}$, \ac{sampool} satisfies \cref{defn:ltag:assoc-aggregation-pooling} and is therefore compatible with \acs{lta}.
To see why this is the case, the associativity and homomorphicity to a weighted aggregation function $\mathcal{A}_{\text{SAM}}$ must be shown.
The associativity of $\mathit{SAM}$ follows directly from its commutativity and the fact that
\begin{gather*}
	\mathit{SAM}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) = \frac{1}{e^{w'} + \sum_{i=j+1}^n e^{w_i}} \left( e^{w'} \underbrace{\mathit{SAM}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^j)}_{y'} + \smashoperator{\sum_{i=j+1}^n} e^{w_i} y_i \right) \\
	\text{for all } j \in [n] \text{ with } w' \coloneqq \ln\sum_{i=1}^{j} e^{w_i} \text{, i.e.\ } \rho_{\mathit{SAM}}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^j) = (y', w')\text{.}
\end{gather*}
The homomorphicity of $\mathit{SAM}$ to a weighted aggregation function $\mathcal{A}_{\text{SAM}}$ follows by choosing $\mathcal{A}_{\text{SAM}}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) \coloneqq {\left(\sum_{i=1}^n w_i\right)}^{-1} \sum_{i=1}^n w_i y_i$ and the trivial homomorphism $\tau_1(y, w) \coloneqq (y, e^w)$, $\tau_2(y) \coloneqq y$ between the algebras of $\mathit{SAM}$ and $\mathcal{A}_{\text{SAM}}$.

This concludes our analysis of the relation between \ac{lta} and \acp{gcnn}.
We saw that spectral convolutions with arbitrary filters are only partially \acs{lta}-like since their constituents generally span entire connected components.
\Acp{gcnn} using vertex neighborhood convolutions on the other hand were shown in \cref{thm:ltag:gcnn-ltag-formulation} to have an \ac{lta} formulation with localized subtree constituents.
The only requirement for this formulation to exist is that an associative pooling layer like $\mean$, $\min$, $\max$ or $\mathit{SAM}$ is used to combine the convolved feature vectors into the final output.

\section{A Novel \acs*{lta}-Inspired \acs*{gcnn} Architecture}%
\label{sec:ltag:wl2gnn}

\subsection{Shortcomings of $k$-\acsp*{gnn}}%
\label{sec:ltag:wl2gnn:kgnn-problems}

\subsection{The 2-\acs*{wl} Convolution Operator}%
\label{sec:ltag:wl2gnn:definition}

\subsection{The Expressive Power of 2-\acs*{wl}-\acsp*{gcnn}}%
\label{sec:ltag:wl2gnn:properties}

\subsection{Efficient Implementation on \acsp*{gpgpu}}%
\label{sec:ltag:wl2gnn:implementation}

>>>>

In the previous chapter an introduction to two separate fields of research was given:
\begin{enumerate*}
	\item \Acf{lta},
	\item \Acf{gcr}
\end{enumerate*}.
In this chapter we will combine them and define an extension of \ac{lta} to the \ac{gcr} problem.
This will be done in three steps:
\begin{enumerate}
	\item We begin with a formal definition of what actually constitutes an \ac{lta} method as opposed to non-\acs{lta} methods.
	\item Using this definition, we will see that an \ac{svm} that uses one of the previously described graph embedding approaches can be interpreted as an \ac{lta} variant under certain conditions.
	\item In the last step we look at the previously described \ac{gcnn} architectures and formalize their relation to \ac{lta}.
\end{enumerate}

@@

\section{\acsp*{svm} with Graph Embeddings as \acs*{lta} Models}%
\label{sec:ltag:formulation:svm}

Based on the general definition of \ac{lta} from the last section, we will now see to what extent the \ac{gcr} methods described in \cref{sec:related:gcr} can be interpreted as \ac{lta} instances.
This section explores the relation between \acp{svm} that use graph embeddings and \ac{lta}.
In \cref{sec:related:gcr:embed,sec:related:gcr:kernel} three different ways to map a given graph $G$ to a vector $\varphi(G) \in \mathbb{R}^d$ were described:
\begin{enumerate*}
	\item Fingerprint embeddings,
	\item skip-gram inspired embeddings and
	\item kernel embeddings
\end{enumerate*}.
One common way to solve the \ac{gcr} problem via those embedding vectors is to train an \ac{svm} on them.
We will now see that \acp{svm} can be interpreted as \ac{lta} models if they are trained on so-called \acp*{sce}.
\begin{defn}
	Given a a multiset $A = \ldblbrace \underbrace{a_1, \dots, a_1}_{\gamma_A(a_1)\text{ times}},\allowbreak \dots, \underbrace{a_n, \dots, a_n}_{\gamma_A(a_n)\text{ times}} \rdblbrace \subseteq D$, the so-called \textit{multiplicity function} $\gamma_A: D \to \mathbb{N}_0$ maps each element of the domain $D$ to its multiplicity in $A$ with $\gamma_A(x) = 0 \Leftrightarrow x \notin A$.
\end{defn}
\begin{defn}\label{defn:ltag:substruct-embedding}
	A graph embedding $\varphi: \mathcal{G} \to \mathbb{N}_0^{d}$ is called a \ac{sce} iff.\ %
	there are decomposition functions $\psi_{\varphi, i}: \mathcal{G} \to \mathcal{P}(\mathcal{G})$ and multiplicity functions $\gamma_{\varphi, i}: \mathcal{G} \to \mathbb{N}_0$ for all embedding components $i \in [d]$ s.t.\ %
	$\forall G \in \mathcal{G}, i \in [d]: \varphi(G)[i] = \sum_{c \in \psi_{\varphi, i}(G)} \gamma_{\varphi, i}(c)$ where $\gamma_{\varphi, i}$ is the multiplicity function of a multiset of the constituents $\psi_{\varphi, i}(G)$.
	We call $\psi_{\varphi, i}$ an \textit{underlying decomposition} of the $i$-th component of the embedding $\varphi$.
	Similarly the joint decomposition $\psi_{\varphi}(G) \coloneqq \bigcup_{i=1}^d \psi_{\varphi,i}(G)$ is an \textit{underlying decomposition} of $\varphi$.
\end{defn}
Intuitively \cref{defn:ltag:substruct-embedding} states that the value of each \ac{sce} component must be derived from the number of constituents produced by some decomposition function where each constituent might be counted multiple times.
Based on this requirement we now proof the main theorem which shows the relation between \acp{svm} and \ac{lta}.

@@

\section{\acsp*{gcnn} as \acs*{lta} Models}%
\label{sec:ltag:formulation:gcnn}

@@

Looking back, an overview of graph convolution and graph pooling layers was given in \cref{sec:related:gcr:nn}.
There are many ways to combine those layers in a concrete \ac{gnn} but we focus on \textit{global pooling \ac{gcnn} architectures} that use the following three steps:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Convolution:}
		First a stack of $T$ convolution layers is applied to the input feature matrix $X \in \mathbb{R}^{n \times d^{(0)}}$ where each row is a feature vector $x_i \in \mathbb{R}^{d^{(0)}}$ for the vertex $v_i \in \mathcal{V}_G$ or the vertex $k$-multiset $s_i \subseteq \mathcal{V}_G$ in case of $k$-\acsp{gnn}.
		The convolution layers produce a convolved feature matrix $Z \in \mathbb{R}^{n \times d^{(T)}}$ where each row is a convolved feature vector $z_i \in \mathbb{R}^{d^{(\mathrm{pool})}}$.
	\item \textbf{Pooling:}
		Then the convolved feature matrix $Z$ is reduced to a single graph feature vector $z_G \in \mathbb{R}^{d^{(\mathrm{pool})}}$ via a pooling layer.
	\item \textbf{\ac{mlp}:}
		Lastly the final output $y_G \in \mathcal{Y}$ is computed by applying a standard \ac{mlp} to the graph feature vector $z_G$.
\end{enumerate}
\[
	h_{\mathrm{\acs*{gcnn}}}(G) \coloneqq \mathrm{\acs*{mlp}}(\mathit{Pool}(\mathit{Conv}(G)))
\]
\Cref{fig:ltag:gcnn-structure} illustrates this three step architecture for vertex neighborhood convolutions like those used by \acp{gcn}~\cite{Kipf2017} or \acp{gin}~\cite{Xu2018}.
Intuitively the convolution layers implicitly define an \ac{lta} decomposition function $\psi$ and a local evaluation function $f$ while the pooling layer roughly corresponds to an \ac{lta} aggregation function $\mathcal{A}$.
This intuition is now formalized.
\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{gfx/graph-lta/gcnn-structure.pdf}
	\caption[Computational steps of a \ac{gcnn} that uses a single vertex neighborhood convolution layer.]{
		Computational steps of a \ac{gcnn} that uses a single vertex neighborhood convolution layer.
		The colored \ac{bfs} trees next to each convolved feature vector $z_i$ show the vertices $v_j$ whose input feature vectors $x_j$ were used to compute $z_i$.
	}\label{fig:ltag:gcnn-structure}
\end{figure}

@@

Based on the decomposition function $\psi: \mathcal{G} \to \mathcal{P}(\mathcal{G})$ and the multi-score evaluation function $f^{*}: \mathcal{G} \to {(\mathcal{Y} \times \mathbb{R}_{\geq 0})}^{*}$ which are implicitly defined by the stack of convolution layers $\mathit{Conv}$ of a given \ac{gcnn}, we now complete our \ac{gcnn} \ac{lta} formulation by defining an evaluation function $f: \mathcal{G} \to \mathcal{Y} \times \mathbb{R}_{\geq 0}$ as well as an aggregation function $\mathcal{A}: {(\mathcal{Y} \times \mathbb{R}_{\geq 0})}^{*} \to \mathcal{Y}$ based on the pooling layer $\mathit{Pool}: \mathbb{R}^{n \times d^{(T)}} \to \mathbb{R}^{d^{(\mathrm{pool})}}$.

@@

	A pooling layer $\mathit{Pool}: \mathbb{R}^{n \times d^{(T)}} \to \mathbb{R}^{d^{(\mathrm{pool})}}$ is called an \textit{associative weighted aggregation layer} iff.

@@

		\item there is a homomorphism $(\tau_{1}: \mathbb{R}^{d^{(T)}} \to (\mathcal{Y} \times \mathbb{R}_{\geq 0}), \tau_{2}: \mathbb{R}^{d^{(\mathrm{pool})}} \to \mathcal{Y})$ from the associative algebra of $\mathit{Pool}$ to that of $\mathcal{A}$

@@

	A \ac{gcnn} of the form $h(G) = \mathit{Pool}(\mathit{Conv}(G))$ has a nontrivial \ac{lta} formulation if $\mathit{Conv}$ is a stack of neighborhood convolution layers and $\mathit{Pool}$ is an associative weighted aggregation layer.

@@

		The \ac{lta} formulation in this proof assumes that the \ac{gcnn}'s output is translated into the target score space via $\tau_{2}: \mathbb{R}^{d^{(\mathrm{pool})}} \to \mathcal{Y}$.
		All this means is that an \acs{lta}-like \ac{gcnn} must produce outputs that are in $\mathcal{Y}$;
		for example if $\mathit{Pool}$ returns one-hot vector encodings of discrete classes, those must be decoded.

@@

The reason for this is that the composition $\mathrm{\acs*{mlp}} \circ \mathit{Pool}$ cannot be guaranteed to be homomorphic to a weighted aggregation function $\mathcal{A}$ due to the universal approximation theorem.

To conclude our analysis of the relation between \ac{lta} and \acp{gcnn}, we check which of the graph pooling layers described in \cref{sec:related:gcr:nn} satisfy \cref{defn:ltag:assoc-aggregation-pooling} are therefore allowed in \acs{lta}-like \acp{gcnn} by \cref{thm:ltag:gcnn-ltag-formulation}.
For simplicity we assume that the target score domain is $\mathcal{Y} = [0, 1]$.
\begin{enumerate}[label={\textbf{\arabic*.}}]
	\item \textbf{Unweighted mean, minimum and maximum pooling:}
		If the last convolution layer has a single output dimension and an activation function with the signature $\sigma: \mathbb{R} \to [0,1]$ (e.g.\ the logistic function), the simple unweighted pooling layers $\mean$, $\min$ and $\max$ trivially satisfy \cref{defn:ltag:assoc-aggregation-pooling} due to their associativity.
	\item \textbf{SortPooling \& \ac{sagpool}:}
		Apart from such unweighted pooling layers we also looked at two state-of-the-art graph pooling layers:
		\begin{enumerate*}[label={\arabic*.}]
			\item SortPooling~\cite{Zhang2018} works by aggregating the top-$p$ convolved vectors $z_i \in \mathbb{R}^{d^{(T)}}$ based on a lexicographic ordering $\prec$,
			\item \ac{sagpool}~\cite{Lee2019} also aggregates the top-$p$ vectors $z_i$ but determines their ordering $\prec$ based on attention weights which are encoded as one component of each $z_i$
		\end{enumerate*}.
		Both of those pooling approaches are not compatible with the \ac{lta} formulation from \cref{thm:ltag:gcnn-ltag-formulation} because top-$p$ aggregation is generally not associative.
		This is shown by the fact that there is no $\rho: \mathbb{R}^{* \times d^{(T)}} \to \mathbb{R}^{ d^{(T)}}$ s.t.\ for all aggregation functions $\mathcal{A}$ and vectors $z_1 \prec z_2$ with $\mathcal{A}(\ldblbrace z_2, z_2 \rdblbrace) = z_2$ (see \cref{defn:ltag:weighted-agg}) it holds that
		\begin{alignat*}{3}
			&& \mathcal{A}(\mathit{top}_{2}(\ldblbrace z_1, z_2, z_2 \rdblbrace)) &= \mathcal{A}(\mathit{top}_{2}(\ldblbrace z_1, \rho(\ldblbrace z_2, z_2 \rdblbrace) \rdblbrace))&& \\
			\Leftrightarrow&& z_2 &= \mathcal{A}(\ldblbrace z_1, \rho(\ldblbrace z_2, z_2 \rdblbrace) \rdblbrace)&&\lightning\text{ if $\mathcal{A} \neq \max$.}
		\end{alignat*}
		From the \ac{lta} perspective the non-associativity of SortPooling and \ac{sagpool} implies that, if they were used in a \ac{gcnn}, it could happen that some subtree evaluations $(y_i, w_i) \in f^{*}(c)$ of a single constituent $c$ are considered by the aggregation function $\mathcal{A}$ while some are ignored depending on the subtree evaluations of other constituents, i.e.\ the constituent evaluation $f(c)$ could then depend on other constituents.
		This is forbidden in our definition of \ac{lta}.
\end{enumerate}

Since neither SortPooling nor \ac{sagpool} are compatible with \ac{lta}, we propose an associative weighted aggregation layer inspired by \ac{sagpool} which uses a softmax-based attention filter instead of a top-$p$ filter.
\begin{defn}
	We define \ac{sampool} as
	\[\mathrm{SAM}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) \coloneqq\frac{1}{\sum_{i=1}^n e^{w_i}} \sum_{i=1}^n e^{w_i} y_i \text{.}\]
\end{defn}
If the stack of convolution layers of a \ac{gcnn} produces vectors $z_i = (y_i, w_i) \in [0, 1] \times \mathbb{R}$, \ac{sampool} satisfies \cref{defn:ltag:assoc-aggregation-pooling} and is therefore compatible with \acs{lta}.
To see why this is the case, the associativity and homomorphicity to a weighted aggregation function $\mathcal{A}_{\mathrm{SAM}}$ must be shown.
The associativity of $\mathrm{SAM}$ follows directly from its commutativity and the fact that
\begin{gather*}
	\mathrm{SAM}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) = \frac{1}{e^{w'} + \sum_{i=j+1}^n e^{w_i}} \left( e^{w'} \underbrace{\mathrm{SAM}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^j)}_{y'} + \smashoperator{\sum_{i=j+1}^n} e^{w_i} y_i \right) \\
	\text{for all } j \in [n] \text{ with } w' \coloneqq \ln\sum_{i=1}^{j} e^{w_i} \text{, i.e.\ } \rho_{\mathrm{SAM}}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^j) = (y', w')\text{.}
\end{gather*}
The homomorphicity of $\mathrm{SAM}$ to a weighted aggregation function $\mathcal{A}_{\mathrm{SAM}}$ follows by choosing $\mathcal{A}_{\mathrm{SAM}}({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) \coloneqq {\left(\sum_{i=1}^n w_i\right)}^{-1} \sum_{i=1}^n w_i y_i$ and the trivial homomorphism $\tau_1(y, w) \coloneqq (y, e^w)$, $\tau_2(y) \coloneqq y$ between the algebras of $\mathrm{SAM}$ and $\mathcal{A}_{\mathrm{SAM}}$.

This concludes our analysis of the relation between \ac{lta} and \acp{gcnn}.
We saw that spectral convolutions with arbitrary filters are only partially \acs{lta}-like since their constituents generally span entire connected components.
\Acp{gcnn} using vertex neighborhood convolutions on the other hand were shown in \cref{thm:ltag:gcnn-ltag-formulation} to have an \ac{lta} formulation with localized subtree constituents.
The only requirement for this formulation to exist is that an associative pooling layer like $\mean$, $\min$, $\max$ or $\mathrm{SAM}$ is used to combine the convolved feature vectors into the final output.
